SciPy 1.0---Fundamental Algorithms for Scientific Computing in Python

### 2018-08-15 scipy-1.0/paper.tex
Formatting

\textbf{History}

\textbf{Project scope}
SciPy aims to provide fundamental algorithms for scientific computing. The
breadth of its scope was derived from the Guide to Available Mathematical
Software classification system (GAMS\cite{boisvert1991guide}). In areas
that move relatively slowly, e.g. linear algebra, SciPy aims to provide
complete coverage. In other areas it aims to provide fundamental building
blocks while interacting well with other packages specialized in that area.
For example, SciPy provides what one expect to find in a
statistics textbook (probability distributions, hypothesis tests, frequency
statistics, correlation functions, and more), while Statsmodels provides
more advanced statistical estimators and inference methods, scikit-learn
covers machine learning, PyMC3, emcee and PyStan cover Bayesian statistics
and probabilistic modelling.

@@

\textbf{Current status (maturity, users)}

@@

\subsection*{Submodule organisation}
\input{subpackages}

@@

\subsection*{Language choices}
Python, Cython, Fortran, C and C++ are the programming languages
used to implement scientific algorithms in the SciPy library. An analysis
of our code base using the \texttt{linguist} library provides a detailed
breakdown as \% composition by programming language in SciPy
(Figure~\ref{fig:linguist}).

@@

\subsection*{API and ABI evolution}
The application programming interface (API) for SciPy consists of
approximately 1500 functions and classes.
Our policy for evolving the API over time is that new functionality can be added,
while removing or changing existing functionality can only be done if the benefits
of that exceeds the (often significant) costs to users, \textit{and} only
after giving clear deprecation warnings to those users for at least one year.
In general, we encourage changes that improve clarity in the API of the
library but strongly discourage breaking backwards compatibility\footnote{
    Backwards compatibility means that new versions of SciPy will keep user
    code written for an older version working.}
given our position near the base of the scientific Python computing stack.

@@

\subsection*{Data structures}
\textbf{cKDTree}
\input{ckdtree}
\textbf{Sparse matrices}

@@

\subsection*{Unified bindings to compiled code}
\subsubsection*{LowLevelCallable}
As of SciPy version 0.19, it is possible for users to wrap low-level
functions in a \texttt{scipy.LowLevelCallable()} object that reduces
the overhead for calling compiled C functions directly from Python.
Supported low-level functions include \texttt{PyCapsule} objects,
ctypes function pointers, and cffi function pointers. The low level
function signature must be consistent with the expectations of the
routine it is passed to. For example, the documentation for
\texttt{scipy.ndimage.generic\_filter} defines two acceptable C callback
function signatures that may be used to produce functions that operate
on each element of image data with low overhead. The C code may be
generated using numba or Cython, for example, as long as the function
call signature matches the specifications. Furthermore, it is even
possible to generate a low-level callback function automatically
from a Cython module using \texttt{scipy.LowLevelCallable.from\_cython}.

@@

\subsection*{Numerical optimization}
\input{scipy-optimize}

\subsection*{Statistical distributions}
The \texttt{scipy.stats} module provides provides more than 100 probability
distributions and a framework to implement additional distributions. Given
an expression for a probability distribution function (pdf),
the framework will calculate moments, cumulative distribution function (cdf),
differential entropy, and more. Those properties are calculated based on
generic code, e.g., numerical integration of the pdf to obtain the cdf.
Where possible, the generic code is replaced by explicit formulas or more
accurate or efficient implementations. For example, most distributions have
an explicit formula for the cdf, so it is used instead of integrating the
pdf. The distributions framework also allows the user to draw random
variates from the distribution. Drawing random variates now accepts a \texttt{random\_state}
parameter, which is either a \texttt{numpy.random.RandomState} object or an
integer used to generate such an object, to provide fine-grained user-level
control over the random state of a workflow.
Currently about 80 continuous univariate distributions, 12 discrete univariate
distributions and 10 multivariate distributions are implemented.
Key recent distributions added to \texttt{scipy.stats} include the histogram-based
distribution in \texttt{scipy.stats.rv\_histogram} and the multinomial
distribution in \texttt{scipy.stats.multinomial} (used, for example, in natural
language processing\cite{Griffiths5228}). \texttt{multinomial} is the first and so 
far the only multivariate discrete distribution available in SciPy.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Polynomial interpolators}
\input{poly}

@@

    \subsubsection*{Test suite}
    The SciPy test suite is orchestrated by a continuous integration matrix
    that includes posix and Windows (32/64-bit) platforms managed by Travis CI and appveyor,
    respectively. Our tests cover Python versions 2.7, 3.4, 3.5, 3.6, and include
    code linting with pyflakes and pycodestyle. There are more than $13000$ unit
    tests in the test suite, which is written for usage with the pytest runner, and
    with a 76.5 \% coverage of approximately $204000$ lines
    of code holding steady for the last six months (Figure~\ref{fig:coverage}).
    Some of the historical components of the code base would still benefit from
    increased test coverage. Documentation for the code is automatically built and hosted 
    by the CircleCI service to facilitate evaluation of documentation changes / integrity.
    Our full test suite also passes with PyPy3, a just-in-time compiled version
    of the Python language.

@@

\textbf{Governance}

@@

\textbf{Roadmap}

@@

# example of the old syntax for handling
# the same scenario
cpdef int func_old(object[int, ndim=3, mode='strided'] arr):
    pass

\end{lstlisting}

@@

\textbf{Community beyond the SciPy library}

\textbf{Maintainers and contributors}
The SciPy project has approximately 15 active maintainers - people who review
others' contributions, and in general do everything needed to ensure that
the software and the project move forward. Maintainers are critical to the
health of the project\cite{eghbal2016}; their skills and efforts largely
determine how fast the project moves forward, and they enable contributions
from a much larger group. Accordingly, the project also has an additional
~100 unique contributors for every 6-monthly release cycle. Anyone with the
interest and skills can become a contributor or maintainer; the SciPy
Developer Guide provides guidance on how to do that.

@@

\textbf{Impact now}
SciPy has a strong developer community and a massive user base. GitHub
traffic metrics report roughly 20,000 unique visitors to the source
website between May 14 and May 27, 2018, with 721 unique copies (``clones")
of the code base over that roughly two-week period. The developer community
consists of 610 unique contributors of source code, with $>19,000$ commits
accepted into the code base (GitHub page data).

@@

\textbf{Future development}.
Future development plans include a major rewrite of the \texttt{scipy.sparse} module, which is already underway\cite{abbasi2018sparse}. This is due to the current interface only supporting two-dimensional objects. Even sub-arrays are two-dimensional, and dimensions greater than two aren't supported. The new package supports dimensions greater than two, and it bases itself around arrays rather than matrices (which is strictly more powerful). It also features improved inter-operability with NumPy and other community packages.

>>>>

\subsection*{History}

\subsection*{Project scope}

SciPy aims to provide fundamental algorithms for scientific computing. The
breadth of its scope was derived from the Guide to Available Mathematical
Software classification system (GAMS\cite{boisvert1991guide}). In areas
that move relatively slowly, e.g. linear algebra, SciPy aims to provide
complete coverage. In other areas it aims to provide fundamental building
blocks while interacting well with other packages specialized in that area.
For example, SciPy provides what one expect to find in a
statistics textbook (probability distributions, hypothesis tests, frequency
statistics, correlation functions, and more), while Statsmodels provides
more advanced statistical estimators and inference methods, scikit-learn
covers machine learning, PyMC3, emcee and PyStan cover Bayesian statistics
and probabilistic modelling.

@@

\subsection*{Current status (maturity, users)}

@@

\subsection*{Submodule organisation}

\input{subpackages}

@@

\subsection*{Language choices}

Python, Cython, Fortran, C and C++ are the programming languages used to
implement scientific algorithms in the SciPy library. An analysis of our code
base using the \texttt{linguist} library provides a detailed breakdown as \%
composition by programming language in SciPy (Figure~\ref{fig:linguist}).

@@

\subsection*{API and ABI evolution}

The application programming interface (API) for SciPy consists of approximately
1500 functions and classes.  Our policy for evolving the API over time is that
new functionality can be added, while removing or changing existing
functionality can only be done if the benefits of that exceeds the (often
significant) costs to users, \textit{and} only after giving clear deprecation
warnings to those users for at least one year.  In general, we encourage
changes that improve clarity in the API of the library but strongly discourage
breaking backwards compatibility\footnote{
    Backwards compatibility means that new versions of SciPy will keep user
    code written for an older version working.}
given our position near the base of the scientific Python computing stack.

@@

\subsection*{Data structures}

\subsubsection*{cKDTree}

\input{ckdtree}

\subsubsection*{Sparse matrices}

@@

\subsection*{Unified bindings to compiled code}

\subsubsection*{LowLevelCallable}

As of SciPy version 0.19, it is possible for users to wrap low-level functions
in a \texttt{scipy.LowLevelCallable()} object that reduces the overhead for
calling compiled C functions directly from Python.  Supported low-level
functions include \texttt{PyCapsule} objects, ctypes function pointers, and
cffi function pointers. The low level function signature must be consistent
with the expectations of the routine it is passed to. For example, the
documentation for \texttt{scipy.ndimage.generic\_filter} defines two acceptable
C callback function signatures that may be used to produce functions that
operate on each element of image data with low overhead. The C code may be
generated using numba or Cython, for example, as long as the function call
signature matches the specifications. Furthermore, it is even possible to
generate a low-level callback function automatically from a Cython module using
\texttt{scipy.LowLevelCallable.from\_cython}.

@@

\subsection*{Numerical optimization}

\input{scipy-optimize}

\subsection*{Statistical distributions}

The \texttt{scipy.stats} module provides provides more than 100 probability
distributions and a framework to implement additional distributions. Given an
expression for a probability distribution function (pdf), the framework will
calculate moments, cumulative distribution function (cdf), differential
entropy, and more. Those properties are calculated based on generic code, e.g.,
numerical integration of the pdf to obtain the cdf.  Where possible, the
generic code is replaced by explicit formulas or more accurate or efficient
implementations. For example, most distributions have an explicit formula for
the cdf, so it is used instead of integrating the pdf. The distributions
framework also allows the user to draw random variates from the distribution.
Drawing random variates now accepts a \texttt{random\_state} parameter, which
is either a \texttt{numpy.random.RandomState} object or an integer used to
generate such an object, to provide fine-grained user-level control over the
random state of a workflow.  Currently about 80 continuous univariate
distributions, 12 discrete univariate distributions and 10 multivariate
distributions are implemented.  Key recent distributions added to
\texttt{scipy.stats} include the histogram-based distribution in
\texttt{scipy.stats.rv\_histogram} and the multinomial distribution in
\texttt{scipy.stats.multinomial} (used, for example, in natural language
processing\cite{Griffiths5228}). \texttt{multinomial} is the first and so far
the only multivariate discrete distribution available in SciPy.

\subsection*{Polynomial interpolators}

\input{poly}

@@

\subsubsection*{Test suite}

The SciPy test suite is orchestrated by a continuous integration matrix that
includes posix and Windows (32/64-bit) platforms managed by Travis CI and
appveyor, respectively. Our tests cover Python versions 2.7, 3.4, 3.5, 3.6, and
include code linting with pyflakes and pycodestyle. There are more than $13000$
unit tests in the test suite, which is written for usage with the pytest
runner, and with a 76.5 \% coverage of approximately $204000$ lines of code
holding steady for the last six months (Figure~\ref{fig:coverage}).  Some of
the historical components of the code base would still benefit from increased
test coverage. Documentation for the code is automatically built and hosted by
the CircleCI service to facilitate evaluation of documentation changes /
integrity.  Our full test suite also passes with PyPy3, a just-in-time compiled
version of the Python language.

@@

\subsection*{Governance}

@@

\subsection*{Roadmap}

@@

# example of the old syntax for handling
# the same scenario
cpdef int func_old(object[int, ndim=3, mode='strided'] arr):
    pass
\end{lstlisting}

@@

\subsection*{Community beyond the SciPy library}

TODO

\subsection*{Maintainers and contributors}

The SciPy project has approximately 15 active maintainers - people who review
others' contributions, and in general do everything needed to ensure that the
software and the project move forward. Maintainers are critical to the health
of the project\cite{eghbal2016}; their skills and efforts largely determine how
fast the project moves forward, and they enable contributions from a much
larger group. Accordingly, the project also has an additional ~100 unique
contributors for every 6-monthly release cycle. Anyone with the interest and
skills can become a contributor or maintainer; the SciPy Developer Guide
provides guidance on how to do that.

@@

\subsection*{Impact now}

SciPy has a strong developer community and a massive user base. GitHub traffic
metrics report roughly 20,000 unique visitors to the source website between May
14 and May 27, 2018, with 721 unique copies (``clones") of the code base over
that roughly two-week period. The developer community consists of 610 unique
contributors of source code, with $>19,000$ commits accepted into the code base
(GitHub page data).

@@

\subsection*{Future development}

Future development plans include a major rewrite of the \texttt{scipy.sparse}
module, which is already underway\cite{abbasi2018sparse}. This is due to the
current interface only supporting two-dimensional objects. Even sub-arrays are
two-dimensional, and dimensions greater than two aren't supported. The new
package supports dimensions greater than two, and it bases itself around arrays
rather than matrices (which is strictly more powerful). It also features
improved inter-operability with NumPy and other community packages.
